---
title: "Line Detection Strategies"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Line Detection Strategies}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)

# Load package
library(houghPrimitives)

# Setup rgl for headless rendering (required for CRAN checks and CI)
if (requireNamespace("rgl", quietly = TRUE)) {
  setup_rgl_knitr()
}
```

## Introduction

Line detection in 3D point clouds is essential for identifying:

- **Infrastructure**: Power lines, cables, transmission lines
- **Building edges**: Roof ridges, wall corners, structural beams
- **Urban features**: Street edges, fence lines, railings
- **Industrial scans**: Pipes, conduits, framework elements

The `detect_lines()` function uses RANSAC-based 3D line fitting with a critical feature: **context-aware detection** via the `in_scene` parameter.

## Two Detection Modes

### Mode 1: Line-Only Scenes (`in_scene = FALSE`)

**Use when:** Your point cloud contains **primarily linear features** (cables, wires, infrastructure edges)

**Behavior:** Aggressive detection with relaxed parameters
- Distance threshold increased by 30%
- Min votes reduced to 70%
- Min length reduced to 80%
- Max iterations increased by 50%

**Example use cases:**
- Power line corridor scans
- Cable/wire detection
- Edge-only extractions

### Mode 2: Complex Scenes (`in_scene = TRUE`) [DEFAULT]

**Use when:** Lines coexist with **other shapes** (planes, cylinders, buildings)

**Behavior:** Stricter detection using exact parameters
- Only detects standalone lines
- Avoids detecting edges of planes as separate lines
- Use AFTER detecting other shapes

**Example use cases:**
- Urban scenes with buildings and standalone cables
- Forestry with terrain and fallen logs
- Industrial scans with structures and conduits

## Basic Usage

```{r eval=FALSE}
library(houghPrimitives)

# Detect lines
result <- detect_lines(
  xyz = point_matrix,       # N x 3 matrix
  distance_threshold = 0.05, # Max distance from line
  min_votes = 30,           # Min points per line
  min_length = 1.0,         # Min line length (meters)
  max_iterations = 500,     # RANSAC iterations
  in_scene = TRUE,          # Context mode
  label_points = FALSE      # Return labels?
)

cat("Detected", result$n_lines, "lines\n")
print(result$lines)
```

## Parameters

### `distance_threshold`

Maximum perpendicular distance from line for a point to be considered part of that line.

```{r eval=FALSE}
# Thin features (cables)
detect_lines(xyz, distance_threshold = 0.02)  # 2cm

# Thicker features (beams)
detect_lines(xyz, distance_threshold = 0.10)  # 10cm

# Noisy data
detect_lines(xyz, distance_threshold = 0.15)  # 15cm
```

**Impact:**
- Too small: Miss valid line points, fragmented lines
- Too large: Include noise, merge nearby parallel lines

### `min_votes`

Minimum number of points required to form a valid line.

```{r eval=FALSE}
# Dense point clouds
detect_lines(xyz, min_votes = 50)

# Sparse or thin features
detect_lines(xyz, min_votes = 20)
```

### `min_length`

Minimum 3D Euclidean length of detected line segment.

```{r eval=FALSE}
# Short segments acceptable
detect_lines(xyz, min_length = 0.5)  # 50cm

# Only long features
detect_lines(xyz, min_length = 5.0)  # 5 meters
```

### `max_iterations`

Number of RANSAC iterations. More iterations = more thorough search but slower.

```{r eval=FALSE}
# Quick scan
detect_lines(xyz, max_iterations = 300)

# Thorough search
detect_lines(xyz, max_iterations = 1000)
```

### `in_scene` **[CRITICAL]**

Controls detection aggressiveness based on scene context.

```{r eval=FALSE}
# Line-only scene (cables, wires)
detect_lines(cables, in_scene = FALSE)  # Aggressive

# Complex scene (buildings + cables)
detect_lines(urban_scan, in_scene = TRUE)  # Strict
```

## Scenario 1: Line-Only Detection

When your point cloud contains **only linear features**:

```{r eval=FALSE}
library(houghPrimitives)

# Example: Cable/power line scan
# Point cloud contains primarily cables with some measurement noise

# Use aggressive mode
result <- detect_lines(
  cable_points,
  distance_threshold = 0.05,
  min_votes = 30,
  min_length = 2.0,
  max_iterations = 1000,
  in_scene = FALSE  # ← Line-only mode
)

cat("Detected", result$n_lines, "cables\n")

# Internally, parameters are relaxed:
# - distance_threshold → 0.065 (30% increase)
# - min_votes → 21 (70% of 30)
# - min_length → 1.6 (80% of 2.0)
# - max_iterations → 1500 (50% increase)
```

**Result:** Finds more lines, suitable when false positives are less concerning than missing features.

## Scenario 2: Lines in Complex Scenes

When lines coexist with planes, cylinders, and other shapes:

```{r eval=FALSE}
# Example: Urban scene with buildings and standalone cables

# STEP 1: Detect dominant features first
planes <- detect_planes(
  urban_points,
  distance_threshold = 0.10,
  min_votes = 150,
  label_points = TRUE
)
cat("Found", planes$n_planes, "planes\n")

# STEP 2: Detect cylinders (poles)
cylinders <- detect_cylinders(
  urban_points,
  min_radius = 0.2,
  max_radius = 0.4,
  min_length = 3.0,
  min_votes = 80,
  label_points = TRUE
)
cat("Found", cylinders$n_cylinders, "poles\n")

# STEP 3: Detect standalone lines from remaining points
lines <- detect_lines(
  urban_points,
  distance_threshold = 0.06,
  min_votes = 40,
  min_length = 2.5,
  max_iterations = 1000,
  in_scene = TRUE,  # ← Strict mode for complex scenes
  label_points = TRUE
)
cat("Found", lines$n_lines, "standalone lines\n")
```

**Key insight:** Detect planes/cylinders first, then use strict mode for lines to avoid false positives.

## Return Values

```{r eval=FALSE}
result <- detect_lines(xyz, label_points = TRUE)

# Number of lines detected
result$n_lines  # Integer

# Line parameters (data.frame)
result$lines
#   line_id: 1, 2, 3...
#   x1, y1, z1: start point
#   x2, y2, z2: end point
#   length: 3D Euclidean length
#   n_points: number of points on this line

# Point labels (if requested)
result$point_labels  # Integer vector: 0 = not a line, 1,2,3... = line ID
```

## Example: Synthetic Lines with Noise

```{r eval=FALSE}
set.seed(42)

# Generate random 3D line
generate_line <- function(x1, y1, z1, x2, y2, z2, n_points = 100, noise = 0.02) {
  t <- runif(n_points, 0, 1)
  x <- x1 + t * (x2 - x1) + rnorm(n_points, sd = noise)
  y <- y1 + t * (y2 - y1) + rnorm(n_points, sd = noise)
  z <- z1 + t * (z2 - z1) + rnorm(n_points, sd = noise)
  cbind(x, y, z)
}

# Create 3 lines
line1 <- generate_line(0, 0, 0, 10, 5, 3, n_points = 120)
line2 <- generate_line(5, 10, 1, 15, 15, 5, n_points = 100)
line3 <- generate_line(2, 2, 4, 8, 12, 6, n_points = 150)

# Add noise
noise <- cbind(runif(100, 0, 20), runif(100, 0, 20), runif(100, 0, 6))

# Combine
all_points <- rbind(line1, line2, line3, noise)

# Detect with aggressive mode (line-only)
result <- detect_lines(
  all_points,
  distance_threshold = 0.05,
  min_votes = 30,
  min_length = 2.0,
  in_scene = FALSE,  # Line-only mode
  label_points = TRUE
)

cat("Detected", result$n_lines, "/ 3 lines\n")
print(result$lines)
```

## Recommended Workflow

### For Line-Only Scenes

```{r eval=FALSE}
# Infrastructure, cables, wire scans
lines <- detect_lines(
  xyz,
  distance_threshold = 0.05,  # Adjust to feature thickness
  min_votes = 30,             # Based on point density
  min_length = 2.0,           # Minimum interesting length
  max_iterations = 1000,      # Thorough search
  in_scene = FALSE            # ← Aggressive mode
)
```

### For Multi-Shape Scenes

```{r eval=FALSE}
# Order matters! Detect in this sequence:

# 1. Planes (walls, ground, roofs)
planes <- detect_planes(xyz, min_votes = 150, label_points = TRUE)

# 2. Spheres (if present)
spheres <- detect_spheres(xyz, min_votes = 50, label_points = TRUE)

# 3. Cylinders (poles, tree boles)
cylinders <- detect_cylinders(xyz, min_votes = 80, label_points = TRUE)

# 4. Lines (standalone cables, edges)
lines <- detect_lines(
  xyz,
  distance_threshold = 0.06,
  min_votes = 40,
  in_scene = TRUE,  # ← Strict mode
  label_points = TRUE
)

# 5. Cubes/cuboids (if needed)
cubes <- detect_cubes(xyz, min_votes = 200, label_points = TRUE)
```

## Visualization

```{r eval=FALSE}
library(lidR)

# Detect lines
result <- detect_lines(xyz, in_scene = FALSE, label_points = TRUE)

# Create LAS with line labels
las <- LAS(data.frame(
  X = xyz[, 1],
  Y = xyz[, 2],
  Z = xyz[, 3],
  LineID = result$point_labels,
  IsLine = as.integer(result$point_labels > 0)
))

# Visualize by line ID
plot(las, color = "LineID", size = 4)

# Or just lines vs non-lines
plot(las, color = "IsLine", size = 4)
```

## 3D Visualization Example

Here's a demonstration of line detection with rgl:

```{r line-3d-viz, echo = TRUE}
library(rgl)

# Generate synthetic lines
set.seed(123)

generate_line <- function(x1, y1, z1, x2, y2, z2, n_points = 100, noise = 0.02) {
  t <- runif(n_points, 0, 1)
  x <- x1 + t * (x2 - x1) + rnorm(n_points, sd = noise)
  y <- y1 + t * (y2 - y1) + rnorm(n_points, sd = noise)
  z <- z1 + t * (z2 - z1) + rnorm(n_points, sd = noise)
  cbind(x, y, z)
}

# Create 3 lines in different orientations
line1 <- generate_line(0, 0, 0, 5, 1, 2, n_points = 120)  # Diagonal
line2 <- generate_line(1, 4, 0, 1, 4, 5, n_points = 100)  # Vertical
line3 <- generate_line(3, 1, 1, 8, 3, 1.5, n_points = 150)  # Horizontal

# Add random noise
noise_pts <- cbind(runif(80, 0, 8), runif(80, 0, 5), runif(80, 0, 5))

# Combine
all_points <- rbind(line1, line2, line3, noise_pts)

# Detect lines
result <- detect_lines(
  all_points,
  distance_threshold = 0.05,
  min_votes = 30,
  min_length = 1.5,
  in_scene = FALSE,
  label_points = TRUE
)

# Visualize results
open3d()
colors <- c("gray80", "steelblue", "coral", "forestgreen", "gold")

for (i in 0:result$n_lines) {
  mask <- result$point_labels == i
  points3d(all_points[mask, 1], all_points[mask, 2], all_points[mask, 3],
           color = colors[i + 1], size = ifelse(i == 0, 3, 6))
}

axes3d()
title3d(sprintf("Detected %d Lines", result$n_lines), "X", "Y", "Z")

# Save snapshot
fig_path <- save_rgl_snapshot("line-detection-3d.png", width = 800, height = 600)
close3d()
```

```{r show-line-3d, echo = FALSE, out.width = "100%", fig.cap = "Line detection results showing three detected lines in different orientations."}
knitr::include_graphics(fig_path)
```

## Troubleshooting

### Problem: No lines detected

**Solutions:**
```{r eval=FALSE}
# 1. Relax parameters
result <- detect_lines(xyz, min_votes = 20, distance_threshold = 0.08)

# 2. Use aggressive mode
result <- detect_lines(xyz, in_scene = FALSE)

# 3. Increase iterations
result <- detect_lines(xyz, max_iterations = 1500)
```

### Problem: Lines fragmented (split into multiple segments)

**Causes:**
- Gaps in point coverage
- `min_votes` too high relative to gaps

**Solutions:**
```{r eval=FALSE}
# Lower min_votes
result <- detect_lines(xyz, min_votes = 20)

# Increase distance_threshold slightly
result <- detect_lines(xyz, distance_threshold = 0.08)

# Post-process to merge collinear segments
# (custom function to merge nearby collinear lines)
```

### Problem: Too many false positives (plane edges detected as lines)

**Solution:**
```{r eval=FALSE}
# Detect planes FIRST
planes <- detect_planes(xyz, label_points = TRUE)

# Then detect lines in strict mode
lines <- detect_lines(xyz, in_scene = TRUE, label_points = TRUE)
```

### Problem: Nearby parallel lines merged

**Causes:**
- `distance_threshold` too large

**Solutions:**
```{r eval=FALSE}
# Reduce distance threshold
result <- detect_lines(xyz, distance_threshold = 0.03)
```

## Advanced: Directional Filtering

Post-process results to filter by direction:

```{r eval=FALSE}
result <- detect_lines(xyz, label_points = TRUE)

# Calculate line directions
lines <- result$lines
dx <- lines$x2 - lines$x1
dy <- lines$y2 - lines$y1
dz <- lines$z2 - lines$z1
len <- sqrt(dx^2 + dy^2 + dz^2)

# Unit direction vectors
ux <- dx / len
uy <- dy / len
uz <- dz / len

# Filter nearly horizontal lines (|uz| < 0.2)
horizontal_mask <- abs(uz) < 0.2
horizontal_lines <- lines[horizontal_mask, ]

cat("Found", nrow(horizontal_lines), "horizontal lines\n")

# Filter nearly vertical lines (|uz| > 0.9)
vertical_mask <- abs(uz) > 0.9
vertical_lines <- lines[vertical_mask, ]

cat("Found", nrow(vertical_lines), "vertical lines\n")
```

## Performance Considerations

```{r eval=FALSE}
# Fast (but may miss some lines)
detect_lines(xyz, max_iterations = 300, min_votes = 50)

# Thorough (slower but more complete)
detect_lines(xyz, max_iterations = 2000, min_votes = 20)

# For large point clouds, consider spatial subdivision
# Detect lines in tiles, then merge results
```

## Comparison: in_scene = TRUE vs FALSE

| Aspect | `in_scene = FALSE` | `in_scene = TRUE` |
|--------|-------------------|------------------|
| **Use Case** | Line-only scenes | Multi-shape scenes |
| **Detection** | Aggressive | Strict |
| **Threshold** | +30% relaxation | As specified |
| **Min Votes** | 70% of value | As specified |
| **Min Length** | 80% of value | As specified |
| **Iterations** | +50% more | As specified |
| **False Positives** | Higher risk | Lower risk |
| **Recall** | Higher | Lower |
| **Precision** | Lower | Higher |

## Best Practices

1. **Choose the right mode**: 
   - Line-only data → `in_scene = FALSE`
   - Complex scenes → `in_scene = TRUE`

2. **Detection order matters**: 
   - In complex scenes, detect planes/cylinders BEFORE lines

3. **Parameter tuning**: 
   - Start with defaults, adjust based on results
   - Balance precision vs recall

4. **Validate results**: 
   - Check line endpoints make geometric sense
   - Verify lengths match expectations

5. **Post-process**: 
   - Merge collinear fragments if needed
   - Filter by direction or location

## Summary

The `detect_lines()` function provides flexible 3D line detection with:

✓ **Context-aware modes** via `in_scene` parameter  
✓ RANSAC-based robust fitting  
✓ Configurable sensitivity parameters  
✓ Point labeling for analysis  

**Key Decision:**
- **Line-only scenes**: Use `in_scene = FALSE` for aggressive detection
- **Complex scenes**: Use `in_scene = TRUE` after detecting other shapes

**Critical workflow tip:** When detecting multiple shape types, always detect lines **after planes** to avoid false positives from planar edges.
