---
title: "Introduction to Shape Classification with houghPrimitives"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to Shape Classification with houghPrimitives}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)

# Load package
library(houghPrimitives)

# Setup rgl for headless rendering (required for CRAN checks and CI)
if (requireNamespace("rgl", quietly = TRUE)) {
  setup_rgl_knitr()
}
```

## Overview

The `houghPrimitives` package provides comprehensive 3D shape detection algorithms for point cloud data using Iterative Hough Transform methods. This vignette introduces the key concepts and basic workflow for classifying shapes in point clouds.

## Supported Shape Types

The package detects six types of 3D primitive shapes:

| Shape | Function | Method | Complexity |
|-------|----------|--------|------------|
| **Spheres** | `detect_spheres()` | Spatial clustering + circularity validation | ★★☆☆☆ |
| **Cylinders** | `detect_cylinders()` | Cross-section analysis | ★★★☆☆ |
| **Planes** | `detect_planes()` | RANSAC plane fitting | ★★☆☆☆ |
| **Lines** | `detect_lines()` | RANSAC 3D line fitting | ★★☆☆☆ |
| **Cubes** | `detect_cubes()` | Orthogonal plane analysis (6 faces) | ★★★★☆ |
| **Cuboids** | `detect_cuboids()` | Multi-plane rectangular fitting | ★★★★☆ |

Additionally, there's a specialized function for forestry applications:

- **Tree Boles**: `detect_tree_boles()` - Vertical cylinder with multi-slice validation ★★★☆☆

## Basic Workflow

### 1. Simple Shape Detection

The simplest approach is to detect one shape type at a time:

```{r eval=FALSE}
library(houghPrimitives)

# Load your point cloud data (xyz matrix with 3 columns)
# xyz <- read.csv("your_data.csv")

# Detect spheres
spheres_result <- detect_spheres(
  xyz,
  min_radius = 0.5,
  max_radius = 2.0,
  min_votes = 50
)

print(spheres_result$n_spheres)
print(spheres_result$spheres)
```

### 2. Point Labeling

Use `label_points = TRUE` to get back which points belong to which shape:

```{r eval=FALSE}
# Detect with point labels
result <- detect_spheres(
  xyz,
  min_radius = 0.5,
  max_radius = 2.0,
  min_votes = 50,
  label_points = TRUE
)

# Points are labeled with shape IDs
# 0 = not part of a sphere
# 1, 2, 3... = sphere ID
point_labels <- result$point_labels
```

### 3. Multi-Shape Classification

For complex scenes with multiple shape types, detect shapes in order from specific to general:

```{r eval=FALSE}
# Recommended detection order:
# 1. Spheres (compact, distinct features)
spheres <- detect_spheres(xyz, min_votes = 50, label_points = TRUE)

# 2. Cylinders (elongated structures)
cylinders <- detect_cylinders(xyz, min_votes = 80, label_points = TRUE)

# 3. Lines (thin features - detect BEFORE planes!)
lines <- detect_lines(xyz, min_votes = 40, in_scene = TRUE, label_points = TRUE)

# 4. Planes (large, dominant features)
planes <- detect_planes(xyz, min_votes = 150, label_points = TRUE)

# 5. Cubes/Cuboids (composite structures)
cubes <- detect_cubes(xyz, min_votes = 200, label_points = TRUE)
```

## Understanding Return Values

Each detection function returns a list with:

- **`n_*`**: Number of shapes detected (e.g., `n_spheres`, `n_cylinders`)
- **Shape parameters**: Data frame with shape-specific attributes
- **`point_labels`** (if requested): Integer vector assigning each point to a shape

### Example: Sphere Detection Returns

```{r eval=FALSE}
result <- detect_spheres(xyz, label_points = TRUE)

# result$n_spheres: integer (e.g., 3)
# result$spheres: data.frame with columns:
#   - sphere_id: 1, 2, 3...
#   - x, y, z: center coordinates
#   - radius: sphere radius
#   - n_points: number of points in this sphere
# result$point_labels: integer vector (length = nrow(xyz))
```

### Example: Cylinder Detection Returns

```{r eval=FALSE}
result <- detect_cylinders(xyz, label_points = TRUE)

# result$n_cylinders: integer
# result$cylinders: data.frame with columns:
#   - cylinder_id: 1, 2, 3...
#   - x1, y1, z1: start point of axis
#   - x2, y2, z2: end point of axis
#   - radius: cylinder radius
#   - n_points: number of points
# result$point_labels: integer vector
```

## Key Parameters

All detection functions share common parameters:

### Geometric Constraints

- **`min_radius`, `max_radius`**: Expected size range (spheres, cylinders)
- **`min_length`, `min_height`**: Minimum extent (cylinders, lines)
- **`distance_threshold`**: Maximum distance from shape for a point to be included

### Detection Sensitivity

- **`min_votes`**: Minimum number of points required to form a valid shape
  - Higher = fewer false positives, may miss small shapes
  - Lower = detect smaller shapes, more false positives

- **`max_iterations`**: Number of RANSAC iterations
  - Higher = more thorough search, slower processing
  - Lower = faster, may miss some shapes

### Point Labeling

- **`label_points`**: Return point-to-shape assignments
  - `TRUE` = get `point_labels` vector in result
  - `FALSE` = only get shape parameters (faster)

## Practical Tips

### 1. Parameter Selection

Start with reasonable defaults based on your data scale:

```{r eval=FALSE}
# For urban/building data (meters)
detect_planes(xyz, distance_threshold = 0.1, min_votes = 100)

# For forestry data (tree detection)
detect_tree_boles(xyz, min_radius = 0.1, max_radius = 0.5, min_height = 1.0)

# For small objects
detect_spheres(xyz, min_radius = 0.01, max_radius = 0.1, min_votes = 30)
```

### 2. Visualization with lidR

Integrate seamlessly with the `lidR` package:

```{r eval=FALSE}
library(lidR)

# Detect shapes
result <- detect_spheres(xyz, label_points = TRUE)

# Create LAS object with labels
las <- LAS(data.frame(
  X = xyz[, 1],
  Y = xyz[, 2],
  Z = xyz[, 3],
  ShapeID = result$point_labels
))

# Visualize
plot(las, color = "ShapeID")
```

### 3. Iterative Refinement

If results are poor, adjust parameters systematically:

```{r eval=FALSE}
# Too many false positives?
# → Increase min_votes, decrease distance_threshold

# Missing shapes?
# → Decrease min_votes, increase max_iterations, increase distance_threshold

# Example progression:
result1 <- detect_spheres(xyz, min_votes = 50)  # Start
result2 <- detect_spheres(xyz, min_votes = 30)  # Detect more
result3 <- detect_spheres(xyz, min_votes = 30, max_iterations = 1500)  # More thorough
```

## Common Use Cases

### Forestry: Tree Detection

```{r eval=FALSE}
# Detect vertical cylinders (tree boles)
trees <- detect_tree_boles(
  forest_points,
  min_radius = 0.05,
  max_radius = 0.5,
  min_height = 1.0,
  vertical_only = TRUE
)

cat("Detected", trees$n_trees, "trees\n")
print(trees$trees)
```

### Urban: Building Analysis

```{r eval=FALSE}
# Detect walls and roofs
planes <- detect_planes(
  building_points,
  distance_threshold = 0.1,
  min_votes = 150
)

# Detect rectangular structures
cuboids <- detect_cuboids(
  building_points,
  min_size = 2.0,
  max_size = 20.0,
  min_votes = 200
)
```

### Infrastructure: Power Lines

```{r eval=FALSE}
# Detect cables/wires
lines <- detect_lines(
  cable_points,
  distance_threshold = 0.05,
  min_votes = 30,
  min_length = 2.0,
  in_scene = FALSE  # Aggressive mode for line-only scenes
)
```

## 3D Visualization Example

Here's a simple multi-shape scene visualization using rgl:

```{r intro-3d-viz, echo = TRUE}
library(rgl)

# Create simple synthetic scene
set.seed(42)

# Generate a cylinder (vertical post)
theta <- seq(0, 2*pi, length.out = 100)
z_vals <- runif(100, 0, 3)
cyl_points <- cbind(
  1 + 0.3 * cos(theta),
  1 + 0.3 * sin(theta),
  z_vals
)

# Generate a plane (ground)
plane_points <- cbind(
  runif(200, 0, 5),
  runif(200, 0, 5),
  rnorm(200, 0, 0.02)
)

# Generate a sphere
phi <- runif(150, 0, 2*pi)
theta2 <- acos(2*runif(150) - 1)
sphere_points <- cbind(
  3.5 + 0.5 * sin(theta2) * cos(phi),
  3.5 + 0.5 * sin(theta2) * sin(phi),
  1.5 + 0.5 * cos(theta2)
)

# Combine all points
all_pts <- rbind(cyl_points, plane_points, sphere_points)

# Visualize
open3d()
points3d(plane_points, color = "gray70", size = 3)
points3d(cyl_points, color = "steelblue", size = 5)
points3d(sphere_points, color = "coral", size = 5)
axes3d()
title3d("Multi-Shape Scene", "X", "Y", "Z")

# Save as static image
fig_path <- save_rgl_snapshot("intro-shapes-3d.png", width = 800, height = 600)
close3d()
```

```{r show-intro-3d, echo = FALSE, out.width = "100%", fig.cap = "Example multi-shape scene with cylinder, plane, and sphere."}
knitr::include_graphics(fig_path)
```

## Next Steps

- See `vignette("sphere-detection")` for detailed sphere detection examples
- See `vignette("line-detection")` for line detection strategies
- See `vignette("multi-shape-classification")` for comprehensive scene analysis
- See `vignette("parameter-tuning")` for advanced parameter optimization

## Summary

The `houghPrimitives` package provides:

✓ Six primitive shape detectors plus specialized tree detection  
✓ Point labeling for shape assignment  
✓ Flexible parameters for different data scales  
✓ Integration with lidR for LiDAR workflows  
✓ Fast C++ implementation  

The key to successful shape classification is:

1. Understanding your data scale and expected shape sizes
2. Choosing appropriate detection order
3. Iteratively tuning parameters
4. Using `label_points = TRUE` for comprehensive analysis
