---
title: "Sphere Detection in Point Clouds"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sphere Detection in Point Clouds}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "80%",
  fig.align = "center"
)

# Load package
library(houghPrimitives)

# Setup rgl for headless rendering (required for CRAN checks and CI)
# This enables 3D visualizations without requiring a display device
if (requireNamespace("rgl", quietly = TRUE)) {
  setup_rgl_knitr()  # Sets rgl.useNULL = TRUE + knitr hooks
}
```

## Introduction

Sphere detection is useful for identifying various spherical objects in point clouds:

- Decorative balls in urban scenes
- Tank ends and dome structures
- Spherical markers or targets
- Tree crowns (approximate)
- Industrial spherical components

The `detect_spheres()` function uses spatial clustering combined with geometric validation to identify spheres in 3D point clouds.

## How Sphere Detection Works

The algorithm follows these steps:

1. **Spatial Clustering**: Groups nearby points using 3D grid-based clustering
2. **Center Estimation**: Iteratively refines sphere center using geometric relationships
3. **Radius Calculation**: Computes radius as mean distance from center to surface points
4. **Validation**: Checks circularity, octant coverage, and radius consistency
5. **Filtering**: Removes candidates that don't meet geometric criteria

### Key Innovation: Iterative Center Refinement

Unlike simple centroid calculation, the algorithm uses the geometric relationship:

$$\text{center} = \text{point} - \text{radius} \times \text{direction}$$

This iterative approach (3 iterations) correctly identifies the sphere center even though surface points don't average to the center.

## Basic Usage

```{r eval=FALSE}
library(houghPrimitives)

# Detect spheres in point cloud
result <- detect_spheres(
  xyz = point_matrix,  # N x 3 matrix (x, y, z)
  min_radius = 0.5,    # Minimum sphere radius
  max_radius = 2.0,    # Maximum sphere radius
  min_votes = 50,      # Minimum points per sphere
  label_points = FALSE # Return point labels?
)

# Check results
cat("Detected", result$n_spheres, "spheres\n")
print(result$spheres)
```

## Parameters

### `min_radius` and `max_radius`

Define the expected size range of spheres to detect.

```{r eval=FALSE}
# Small objects (cm scale)
detect_spheres(xyz, min_radius = 0.01, max_radius = 0.1)

# Medium objects (dm to m scale)
detect_spheres(xyz, min_radius = 0.5, max_radius = 2.0)

# Large structures (m scale)
detect_spheres(xyz, min_radius = 2.0, max_radius = 10.0)
```

**Tips:**
- Set range based on expected object sizes in your data
- Too narrow: miss valid spheres
- Too wide: longer processing time, more false positives

### `min_votes`

Minimum number of points required to form a valid sphere.

```{r eval=FALSE}
# Dense point cloud → higher threshold
detect_spheres(xyz, min_votes = 100)  # Strict

# Sparse point cloud → lower threshold  
detect_spheres(xyz, min_votes = 30)   # Permissive

# Balance quality vs completeness
detect_spheres(xyz, min_votes = 50)   # Moderate
```

**Impact:**
- Higher `min_votes`: Fewer false positives, may miss small/partial spheres
- Lower `min_votes`: Detect more spheres, risk of noise clusters

### `label_points`

Whether to return point-to-sphere assignments.

```{r eval=FALSE}
# Get point labels
result <- detect_spheres(xyz, label_points = TRUE)

# Labels are 0 (not a sphere) or 1, 2, 3... (sphere ID)
labels <- result$point_labels

# Count points per sphere
table(labels[labels > 0])
```

## Return Values

The function returns a list with:

```{r eval=FALSE}
result <- detect_spheres(xyz, label_points = TRUE)

# Number of spheres detected
result$n_spheres  # Integer

# Sphere parameters (data.frame)
result$spheres
#   sphere_id: 1, 2, 3...
#   x, y, z: center coordinates
#   radius: sphere radius
#   n_points: number of points in this sphere

# Point labels (if requested)
result$point_labels  # Integer vector, length = nrow(xyz)
```

## Example: Synthetic Spheres with Noise

```{r eval=FALSE}
library(houghPrimitives)
set.seed(42)

# Function to generate sphere surface points
generate_sphere <- function(cx, cy, cz, radius, n_points = 200) {
  # Uniform distribution on sphere surface
  theta <- runif(n_points, 0, 2 * pi)
  phi <- acos(2 * runif(n_points) - 1)
  
  x <- cx + radius * sin(phi) * cos(theta)
  y <- cy + radius * sin(phi) * sin(theta)
  z <- cz + radius * cos(phi)
  
  cbind(x, y, z)
}

# Generate 3 spheres
sphere1 <- generate_sphere(5, 5, 3, radius = 1.2, n_points = 300)
sphere2 <- generate_sphere(12, 8, 2.5, radius = 0.8, n_points = 250)
sphere3 <- generate_sphere(8, 15, 4, radius = 1.5, n_points = 400)

# Add uniform noise
noise <- cbind(
  runif(300, 0, 20),
  runif(300, 0, 20),
  runif(300, 0, 6)
)

# Combine
all_points <- rbind(sphere1, sphere2, sphere3, noise)

# Detect spheres
result <- detect_spheres(
  all_points,
  min_radius = 0.5,
  max_radius = 2.0,
  min_votes = 50,
  label_points = TRUE
)

cat("Detected", result$n_spheres, "spheres\n")
print(result$spheres)

# Calculate accuracy
n_sphere_points <- nrow(sphere1) + nrow(sphere2) + nrow(sphere3)
labels <- result$point_labels

true_positives <- sum(labels[1:n_sphere_points] > 0)
precision <- true_positives / sum(labels > 0)
recall <- true_positives / n_sphere_points

cat(sprintf("Precision: %.2f%%\n", precision * 100))
cat(sprintf("Recall: %.2f%%\n", recall * 100))
```

## 3D Visualization with rgl

Visualize detection results in interactive 3D:

```{r sphere-detection-3d, eval = requireNamespace("rgl", quietly = TRUE)}
library(rgl)
library(houghPrimitives)

# Generate synthetic spheres with noise (same as above example)
set.seed(42)

generate_sphere <- function(cx, cy, cz, radius, n_points = 200) {
  theta <- runif(n_points, 0, 2 * pi)
  phi <- acos(2 * runif(n_points) - 1)
  x <- cx + radius * sin(phi) * cos(theta)
  y <- cy + radius * sin(phi) * sin(theta)
  z <- cz + radius * cos(phi)
  cbind(x, y, z)
}

sphere1 <- generate_sphere(5, 5, 3, radius = 1.2, n_points = 300)
sphere2 <- generate_sphere(12, 8, 2.5, radius = 0.8, n_points = 250)
sphere3 <- generate_sphere(8, 15, 4, radius = 1.5, n_points = 400)

noise <- cbind(runif(300, 0, 20), runif(300, 0, 20), runif(300, 0, 6))
all_points <- rbind(sphere1, sphere2, sphere3, noise)

# Detect spheres
result <- detect_spheres(all_points, min_radius = 0.5, max_radius = 2.0,
                        min_votes = 50, label_points = TRUE)

# Create 3D visualization
open3d()

# Plot undetected points (gray noise)
noise_mask <- result$point_labels == 0
if (any(noise_mask)) {
  points3d(all_points[noise_mask, 1], 
           all_points[noise_mask, 2], 
           all_points[noise_mask, 3],
           color = "gray", size = 3)
}

# Plot detected sphere points (colored by sphere ID)
if (result$n_spheres > 0) {
  colors <- c("steelblue", "coral", "forestgreen", "gold", "purple")
  
  for (i in 1:result$n_spheres) {
    sphere_mask <- result$point_labels == i
    
    # Points on sphere surface
    points3d(all_points[sphere_mask, 1], 
             all_points[sphere_mask, 2], 
             all_points[sphere_mask, 3],
             color = colors[i], size = 5)
    
    # Overlay detected sphere (semi-transparent)
    sphere_params <- result$spheres[i, ]
    spheres3d(sphere_params$x, 
              sphere_params$y, 
              sphere_params$z,
              radius = sphere_params$radius,
              color = colors[i], alpha = 0.2)
  }
}

axes3d()
title3d(sprintf("Detected %d Spheres", result$n_spheres), 
        "X (m)", "Y (m)", "Z (m)")

# Save static PNG snapshot (goes to man/figures/ for pkgdown)
fig_path <- save_rgl_snapshot("sphere-detection-3d.png", 
                              width = 900, height = 700)
close3d()
```

Display the static figure:

```{r show-sphere-3d, echo = FALSE, out.width = "100%", fig.cap = "3D visualization of sphere detection results. Detected spheres shown with semi-transparent overlays."}
knitr::include_graphics(fig_path)
```

For HTML output, you can also create an interactive widget:

```{r sphere-widget, eval = knitr::is_html_output() && requireNamespace("rgl", quietly = TRUE), echo = FALSE}
# Recreate scene for interactive widget
open3d()

noise_mask <- result$point_labels == 0
if (any(noise_mask)) {
  points3d(all_points[noise_mask, 1], 
           all_points[noise_mask, 2], 
           all_points[noise_mask, 3],
           color = "gray", size = 3)
}

if (result$n_spheres > 0) {
  colors <- c("steelblue", "coral", "forestgreen", "gold", "purple")
  for (i in 1:result$n_spheres) {
    sphere_mask <- result$point_labels == i
    points3d(all_points[sphere_mask, 1], 
             all_points[sphere_mask, 2], 
             all_points[sphere_mask, 3],
             color = colors[i], size = 5)
    sphere_params <- result$spheres[i, ]
    spheres3d(sphere_params$x, sphere_params$y, sphere_params$z,
              radius = sphere_params$radius,
              color = colors[i], alpha = 0.2)
  }
}

axes3d()
title3d(sprintf("Detected %d Spheres (Interactive)", result$n_spheres), 
        "X", "Y", "Z")

widget <- create_rgl_widget(width = 800, height = 600)
close3d()

widget
```

## Example: Spheres in Complex Scene

When spheres coexist with other shapes, detect them early in your workflow:

```{r eval=FALSE}
# Load point cloud with multiple shape types
# xyz <- load_your_data()

# STEP 1: Detect spheres first (they're distinct)
spheres <- detect_spheres(
  xyz,
  min_radius = 0.5,
  max_radius = 2.0,
  min_votes = 80,
  label_points = TRUE
)

cat("Found", spheres$n_spheres, "spheres\n")

# STEP 2: Detect other shapes
cylinders <- detect_cylinders(xyz, min_votes = 100, label_points = TRUE)
planes <- detect_planes(xyz, min_votes = 150, label_points = TRUE)

# Combine labels
combined_labels <- rep(0, nrow(xyz))
combined_labels[spheres$point_labels > 0] <- 1  # Sphere
combined_labels[cylinders$point_labels > 0] <- 2  # Cylinder
combined_labels[planes$point_labels > 0] <- 3  # Plane
```

## Visualization with lidR

```{r eval=FALSE}
library(lidR)

# Detect spheres
result <- detect_spheres(xyz, min_votes = 50, label_points = TRUE)

# Create LAS object
las <- LAS(data.frame(
  X = xyz[, 1],
  Y = xyz[, 2],
  Z = xyz[, 3],
  SphereID = result$point_labels,
  IsSphere = as.integer(result$point_labels > 0)
))

# Visualize by sphere ID
plot(las, color = "SphereID", size = 3)

# Visualize sphere vs non-sphere
plot(las, color = "IsSphere", size = 3)
```

## Detection Quality Metrics

### Precision and Recall

When you have ground truth:

```{r eval=FALSE}
# Assume first N points are sphere points, rest is noise
n_true_spheres <- 950
labels <- result$point_labels

# True positives: sphere points correctly classified
tp <- sum(labels[1:n_true_spheres] > 0)

# False positives: noise classified as spheres
fp <- sum(labels[(n_true_spheres + 1):length(labels)] > 0)

# False negatives: sphere points missed
fn <- sum(labels[1:n_true_spheres] == 0)

precision <- tp / (tp + fp)
recall <- tp / (tp + fn)
f1_score <- 2 * (precision * recall) / (precision + recall)

cat(sprintf("Precision: %.2f%%\n", precision * 100))
cat(sprintf("Recall: %.2f%%\n", recall * 100))
cat(sprintf("F1 Score: %.3f\n", f1_score))
```

### Geometric Accuracy

Compare detected parameters to ground truth:

```{r eval=FALSE}
# Ground truth
true_center <- c(5, 5, 3)
true_radius <- 1.2

# Detected
detected <- result$spheres[1, ]

# Center error (Euclidean distance)
center_error <- sqrt(
  (detected$x - true_center[1])^2 +
  (detected$y - true_center[2])^2 +
  (detected$z - true_center[3])^2
)

# Radius error
radius_error <- abs(detected$radius - true_radius)

cat(sprintf("Center error: %.3f meters\n", center_error))
cat(sprintf("Radius error: %.3f meters (%.1f%%)\n", 
            radius_error, 100 * radius_error / true_radius))
```

## Troubleshooting

### Problem: No spheres detected

**Possible causes:**
- `min_votes` too high
- Radius range (`min_radius`, `max_radius`) doesn't match data
- Point density too low

**Solutions:**
```{r eval=FALSE}
# Try more permissive parameters
result <- detect_spheres(
  xyz,
  min_radius = 0.1,      # Wider range
  max_radius = 5.0,
  min_votes = 20         # Lower threshold
)

# Check point cloud summary
summary_stats <- summarize_points(xyz)
print(summary_stats)
```

### Problem: Too many false positives

**Possible causes:**
- `min_votes` too low
- Noise clusters being detected as spheres

**Solutions:**
```{r eval=FALSE}
# Increase strictness
result <- detect_spheres(
  xyz,
  min_votes = 100,        # Higher threshold
  min_radius = 0.8,       # Tighter range
  max_radius = 1.5
)
```

### Problem: Spheres merged or split incorrectly

**Possible causes:**
- Spheres too close together
- Similar radii causing clustering issues

**Solutions:**
- Pre-segment point cloud by spatial regions
- Detect spheres in separate regions
- Post-process results to merge/split as needed

## Advanced: Partial Sphere Detection

The algorithm can detect partial spheres (e.g., hemisphere visible from scanning position):

```{r eval=FALSE}
# Generate hemisphere (only upper half)
generate_hemisphere <- function(cx, cy, cz, radius, n_points = 150) {
  theta <- runif(n_points, 0, 2 * pi)
  phi <- runif(n_points, 0, pi/2)  # Only upper hemisphere
  
  x <- cx + radius * sin(phi) * cos(theta)
  y <- cy + radius * sin(phi) * sin(theta)
  z <- cz + radius * cos(phi)
  
  cbind(x, y, z)
}

hemisphere <- generate_hemisphere(10, 10, 2, radius = 1.0)

# Detection works but requires fewer octants
result <- detect_spheres(hemisphere, min_votes = 40)
# Algorithm uses relaxed octant coverage (3 of 8 octants)
```

## Best Practices

1. **Know your data scale**: Set `min_radius` and `max_radius` appropriately
2. **Start conservative**: Begin with higher `min_votes`, then relax if needed
3. **Detect spheres early**: In multi-shape scenes, detect spheres before planes
4. **Use point labels**: Enable `label_points = TRUE` for analysis and visualization
5. **Validate results**: Check geometric accuracy against known objects
6. **Iterate parameters**: Adjust based on precision/recall metrics

## Summary

The `detect_spheres()` function provides robust sphere detection using:

✓ Iterative center refinement (not simple centroid)  
✓ Geometric validation (octant coverage, radius consistency)  
✓ Flexible parameters for different scales and densities  
✓ Support for partial spheres  
✓ Point labeling for detailed analysis  

Key parameters:
- `min_radius`, `max_radius`: Define size range
- `min_votes`: Control sensitivity/specificity tradeoff
- `label_points`: Get point-to-sphere assignments

For complex scenes with multiple shape types, detect spheres first, then proceed to other shapes.
